---
title: "IUCN Redlist"
author: "Sachiko Lamen"
date: "3/10/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(sf)
library(raster)
library(rnaturalearth)
library(tidyverse)
library(here)
library(jsonlite) # to access API
```

### Accessing IUCN API
```{r}
api_key <-Sys.getenv('IUCN_KEY')
```


```{r}
### append the endpoint to the domain name: 
domain_name <- 'http://apiv3.iucnredlist.org'
version_end <- 'api/v3/version'

version_url <- file.path(domain_name, version_end)
api_version <- jsonlite::fromJSON(version_url)
api_version$version
```
### Get a count of how many species have been assessed

spp count endpoint: `/api/v3/speciescount?token='YOUR TOKEN'`
```{r}
count_stem <- 'api/v3/speciescount?token=%s' 
  ### a format string for sprintf - %s means replace this with character string var;
  ### lots of numeric formats supported, 
  ### e.g., sprintf('%f', pi) vs sprintf('%.3f', pi)
count_end <- sprintf(count_stem, api_key)
count_url <- file.path(domain_name, count_end)

spp_count <- jsonlite::fromJSON(count_url)
spp_count$count
```

### Get a page of results

page endpoint: `/api/v3/species/page/:page_number?token='YOUR TOKEN'`
```{r}
page_stem <- 'api/v3/species/page/%s?token=%s'
page_end <- sprintf(page_stem, 1, api_key) ### multiple variables into format string
page_url <- file.path(domain_name, page_end)

spp_page1 <- fromJSON(page_url)

### convert extinction risk into factor and plot counts
spp_df <- spp_page1$result %>%
  mutate(category = factor(category, levels = c('LC', 'NT', 'VU', 'EN', 'CR', 'EX'))) %>%
  filter(!is.na(category))

ggplot(spp_df) +
  geom_bar(aes(x = category, fill = category)) +
  scale_fill_brewer(palette = 'RdYlGn', direction = -1)
```

### Get current and historic assessment info for a species

historic assessments endpoint: `/api/v3/species/history/name/:name?token='YOUR TOKEN'`

```{r}
hist_stem <- 'api/v3/species/history/name/%s?token=%s'
spp <- 'Dermochelys coriacea'
hist_end <- sprintf(hist_stem, spp, api_key)
hist_url <- file.path(domain_name, hist_end)

spp_hist <- fromJSON(hist_url)

spp_hist_df <- spp_hist$result
```

### Pull threats and narrative, and extract information on gear types

Threats: `/api/v3/threats/species/name/:name?token='YOUR TOKEN'`
Narratives: `/api/v3/species/narrative/:name?token='YOUR TOKEN'`
```{r}
threats_stem <- 'api/v3/threats/species/name/%s?token=%s'
thr_url <- file.path(domain_name, sprintf(threats_stem, spp, api_key))
spp_thr <- fromJSON(thr_url)$result

narratives_stem <- 'api/v3/species/narrative/%s?token=%s'
narr_url <- file.path(domain_name, sprintf(narratives_stem, spp, api_key))
spp_narr <- fromJSON(narr_url)$result

spp_narr$threats
```

## MAKE A MAP
- Read in IUCN and AquaMaps data

```{r}
iucn_spp_info <- read_csv(here('data/iucn_marine_spp_info_2021-3.csv'))
iucn_spp_info$cat %>% unique()

cell_ids <- read_csv(here('data/am_latlong.csv'))
spp_cells <- read_csv(here('data/am_spp_cells.csv'))
spp_ids <- read_csv(here('data/am_spp_ids.csv'))


spp_risk_cells <- spp_cells %>%
  inner_join(cell_ids, by = 'loiczid') %>%
  inner_join(spp_ids, by = 'am_sid') %>%
  inner_join(iucn_spp_info, by = c('binomial' = 'sciname')) # link dataframes on column names that dont match but have the same contents :)

threatened_pct_cells <- spp_risk_cells %>%
  filter(prob >= 0.5) %>%
  mutate(threatened = (tolower(cat) %in% c('vu', 'en', 'cr'))) %>% # is this lowercase categry in this vector -- returns column threatened with values of true or false
  filter(cat != 'dd') %>%
  group_by(lon, lat) %>%
  summarize(pct_threatened = sum(threatened) / n())
```

### Convert cells into a raster object

If we wished to do spatial analysis with this lat-long map of percent threatened species, we can convert our grid to a raster.  Don't forget to tell it our CRS, which since we are using lat-long data, we can use WGS84, EPSG code 4326.

```{r}
spp_risk_rast <- rasterFromXYZ(threatened_pct_cells)
crs(spp_risk_rast) <- 4326

plot(spp_risk_rast)
```

### Plot with ggplot

`ggplot2::geom_raster` does not actually plot `raster` format objects, but gridded data in a dataframe with values that can be assigned to x and y.  If we had data with a native `raster` format, we could convert to a `data.frame` (the reverse of `rasterFromXYZ`) using `as.data.frame(r, xy = TRUE)` - but we already have that in the form of threatened_pct_cells, so let's see if they are different!  

```{r}
spp_risk_rast_df <- spp_risk_rast %>%
  as.data.frame(xy = TRUE)
### Why does this have different length than threatened_pct_cells?

ggplot(threatened_pct_cells) +
  geom_raster(aes(x = lon, y = lat, fill = pct_threatened)) +
  scale_fill_viridis_c() +
  coord_sf()
```

### Let's finalize our map by overlaying a continent shape

We will use vector data from Natural Earth, using the `rnaturalearth` package.  To get higher resolution vector data of global countries, install the `rnaturalearthdata` and `rnaturalearthhires` packages too.

Note, we can access individual countries, or continents, or states; but we probably also want to crop the resulting geometry down to just our area of focus.  Because we made a raster object from our grid of data, we can use that as our crop bounding box.

```{r}
land_sf <- rnaturalearth::ne_countries(scale = 50, ### start with 110
                                       country = c('united states of america', 'mexico'),
                                       returnclass = 'sf')
# plot(land_sf %>% select(geometry))
# st_crs(land_sf)

land_sf_cropped <- land_sf %>%
  st_crop(spp_risk_rast)

ggplot() +
  geom_raster(data = threatened_pct_cells, aes(x = lon, y = lat, fill = pct_threatened)) +
  geom_sf(data = land_sf_cropped, fill = 'grey80', color = 'grey40', alpha = .5) +
  scale_fill_viridis_c() +
  coord_sf() +
  theme_void()

```



